\Title

### 2022-09-22 Grimaldi-etal-BiolCybernetics.tex
up
import from 2022-07-11_Grimaldi-etal-ICIP.tex

@@

%!TeX encoding = UTF-8 Unicode

@@

Learning hetero-synaptic delays for fast motion detection in a single layer of spiking neurons

@@

%%%% Standard Packages
%%<additional latex packages if required can be included here>
\usepackage[lofdepth,lotdepth,position=top]{subfig}
\usepackage{tabularx}
%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\usepackage%[disable]	% uncomment to hide all the notes
%	{todonotes}
%\newcommand{\noteAG}[1]{\todo[author=\textbf{Antoine}, inline]{#1}}
\usepackage{soul}
\usepackage{color}

\DeclareRobustCommand{\note}[1]{{\sethlcolor{yellow}\hl{#1}}}
%\newcommand{\noteLP}[1]{\todo[author=\textbf{Laurent}, inline]{#1}}
%\newcommand{\modif}[1]{\todo[author=\textbf{Proposition}, inline]{#1}}

@@

We can now define an event-based input to the model as a set on incoming spikes reaching the presynaptic neurons at specified times, like the sensory signal representing the output of an event-based camera. It thus forms a discrete stream of events, which can be formalized as an ordered set of addresses and timestamps: $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event $\event_\arank$. Each event has a time of occurrence $\timev_\arank$. The corresponding set of input presynaptic spikes $\event$ will be integrated by this synaptic set and notably by the respective delays, which will multiplex in time all possible motifs. For each time $\timev$ the integration of $\event$ is defined by a list of weights $\activeweights^\postsynaddr$ linked to the synapses that match a precise spatio-temporal motif as input:
$
\activeweights^\postsynaddr (t) = 
\{ \synapticweight_\ranksyn^\postsynaddr  \vert 
\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr \; \text{and} \; \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr \}_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} 
\}
$.
\note{finish by saying that it is effortless in biology but difficult on conventional computers}

@@

The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%
In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
%
\subsection{Task definition: fast motion detection}
%%%-----------------------------------------------------------------
\note{c'est la meme expe faite par baudot et modelisee par kremkow}
Let us now define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the field of vision. In animals with a fovea, this is particularly useful as it allows to move the area with the highest density of photoreceptors in the environment, for example at a point of interest. For example, saccades are rapid movements of the eyeball that reposition the center of vision. In humans, these are very frequent (on average one every 3 per second~\citep{dandekar_neural_2012}). They are produced very rapidly (about $80~\ms$) and display a whole range of speeds. At a more microscopic scale, the human gaze moves with an incessant drift similar to a Brownian-like trajectory~\citep{poletti_head-eye_2015}. To maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D motions in polar coordinates. Based on the distribution of biological motions, we simplified it by selecting a set of eye movements as the Cartesian product of N-V-phi=8 linearly spaced motion directions and N-Y-velocity=7 different velocities (see inset of \seeFig{motion_task}-(a)). Note that the velocities are sampled on a geometric scale between $1/4$ and $v-speed_base = 4$. Next, we define one trajectory of gaze as segments whose duration is drawn from a Poisson distribution with an average block length = $20~\ms$ - similar to a Lévy flight~\citep{}. Finally, the trajectory is integrated by assuming 1/ that velocities are uniformly and independently sampled from the set of different motion-sets and 2/ that motion is uniform during a time segment. The resulting instances yield trajectories qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(a)).

>>>>

% !TEX encoding = UTF-8 Unicode

@@

@@

Learning hetero-synaptic delays in a layer of spiking neurons for fast motion detection

@@

\usepackage[T1]{fontenc}
%\usepackage{textcomp}
\usepackage[utf8]{inputenc}

%%%% Standard Packages
%%<additional latex packages if required can be included here>
%\usepackage[lofdepth,lotdepth,position=top]{subfig}
%\usepackage{tabularx}
%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\usepackage%[disable]	% uncomment to hide all the notes
%	{todonotes}
%\newcommand{\noteAG}[1]{\todo[author=\textbf{Antoine}, inline]{#1}}
\usepackage{soulutf8}
\usepackage{color}

%\DeclareRobustCommand
\newcommand{\note}[1]{{\sethlcolor{yellow}\hl{#1}}}
%\newcommand{\noteLP}[1]{\todo[author=\textbf{Laurent}, inline]{#1}}
%\newcommand{\modif}[1]{\todo[author=\textbf{Proposition}, inline]{#1}}

@@

We can now define an event-based input to the model as a set on incoming spikes reaching the presynaptic neurons at specified times, like the sensory signal representing the output of an event-based camera. It thus forms a discrete stream of events, which can be formalized as an ordered set of addresses and timestamps: $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event $\event_\arank$. Each event has a time of occurrence $\timev_\arank$. The corresponding set of input presynaptic spikes $\event$ will be integrated by this synaptic set and notably by the respective delays, which will multiplex in time all possible motifs. For each time $\timev$ the integration of $\event$ is defined by a list of weights $\activeweights^\postsynaddr$ linked to the synapses that match a precise spatio-temporal motif as input:
$
\activeweights^\postsynaddr (t) = 
\{ \synapticweight_\ranksyn^\postsynaddr  \vert 
\presynaddr_\arank = 
 \presynaddr_\ranksyn^\postsynaddr \; \text{and} \; \timev=\timev_\arank + 
\synapticdelay_\ranksyn^\postsynaddr \}_{ \arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]} 
\}
$.
\note{finish by saying that it is effortless in biology but difficult on conventional computers}
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=0.90\linewidth]{figures/figure_1a.pdf}
    \includegraphics[width=0.90\linewidth]{figures/figure_1b.pdf}
    \includegraphics[width=0.90\linewidth]{figures/figure_1c.pdf}
	    \caption{The hetero-synaptic delays model applied on synthetic data. {\bf (a)}~The afferent information consists of the repeated occurrence of groups of precise motifs of spikes that we call ``polychronous groups'' (PGs). We highlight them by different colors, an information hidden to a detection model. {\bf (b)}~The model is defined as an assembly of neurons (here for $4$ PGs) each defined by a set of different synapses defined by weights (increasing with the radius of the black dots) at each different delay. Note that each afferent may be connected with multiple weights at different delays. The propagation of the afferent information through these delay may generate at each time step a synchronous pattern on a subset of synapses. %These PG patterns were efficiently learned by a logistic regression.
	    {\bf (c)}~The output of the model provides with the predicted probability of occurrence of each PG pattern at any time, which may be used to generate a spike as a Bernoulli trial, providing in this particular case with an exact identification of PGs occurrences. }
    \label{fig:model}
\end{figure}

% 
The sensory signal representing the output of an event-based camera forms a discrete stream of events, which can be formalized as an ordered set of addresses and timestamps: $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event $\event_\arank$. 
This event has a time of occurrence $\timev_\arank$  and an associated address, which is typically in the form $\presynaddr_\arank=(x_\arank, y_\arank, \polev_\arank)$.
This defines a presynaptic address space $\presynaddrspace = [1, \Nx]\times[1, \Ny] \times [1, \Npol] \subset \mathbb{N}^3$ where $(\Nx, \Ny)$ is the size of the sensor in pixels and $\Npol$ is the number of polarities  ($\Npol=2$ for ON and OFF polarities).

We may now define a layer of neurons $\postsynaddr \in \postsynaddrspace$ by first describing how each neuron connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses, and each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then not only described by synaptic weights connecting to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. For each neuron $\postsynaddr$, we define a set of $\Nsyn^\postsynaddr$ synapses, as  $\synapse^\postsynaddr = \{(\presynaddr^\postsynaddr_\ranksyn, \synapticweight^\postsynaddr_\ranksyn, \synapticdelay^\postsynaddr_\ranksyn)\}_{\ranksyn \in [1,\Nsyn^\postsynaddr]}$, where each synapse $\synapse^\postsynaddr_\ranksyn$ is associated to a weight $\synapticweight^\postsynaddr_\ranksyn$, a delay $\synapticdelay^\postsynaddr_\ranksyn$ and a presynaptic address $\presynaddr^\postsynaddr_\ranksyn$. Note that a neuron may contact an afferent neuron with multiple, different delays.

The corresponding input presynaptic spikes $\event$ will be integrated by this synaptic set and notably by the respective delays, which will multiplex in time all possible patterns.
For each time $\timev$ the integration of $\event$ is defined by a list of weights $\activeweights^\postsynaddr$ linked to the synapses that match a precise spatio-temporal motif as input:
$
\activeweights^\postsynaddr(t) = \{\synapticweight_\ranksyn^\postsynaddr | \presynaddr_\arank = \presynaddr_\ranksyn^\postsynaddr \; \text{and} \; \timev=\timev_\arank + \synapticdelay_\ranksyn^\postsynaddr\}_{\arank \in [1,\numevent]\text{, }\ranksyn \in [1,\Nsyn^\postsynaddr]}
$.
%
The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\cite{Grimaldi22pami}, in analogy to a spiking Winner-Take-All network~\cite{nessler2013bayesian}. It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; | \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%

@@

The activation function of our spiking neuron is a softmax function implementing a form of  Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:
$
Pr(k=\postsynaddr \; \vert \; \timev) =
\frac 1 Z
{\exp  (\mathcal{C}^\postsynaddr(\timev) +\bias^\postsynaddr) }
$ 
where $\mathcal{C}^\postsynaddr(\timev) = \sum
\activeweights^\postsynaddr(t)
$ is the sum of the synaptic weights and $\bias^\postsynaddr$ is the bias linked to neuron $\postsynaddr$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike.
%In our MLR model with $\Nclass=\Nspeed$ classes, a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
In our MLR model with $\Nclass=\Nspeed$ classes, 
a probability value is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. 
The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the hetero-synaptic delays model. 
From the perspective of simulating such event-based computations on standard chips, it is useful to transform this sparse representation into a dense representations. As such, we may first write any event-based input as the boolean matrix $A \in \{0, 1 \}^\presynaddrspace$. In this simplified model, we will consider that hetero-synaptic delays are limited in range such that the synaptic set can be represented by the dense matrix $\kernel^\postsynaddr$ giving for each neuron $\postsynaddr$ the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in [1,\Nsyn^\postsynaddr]}, \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) = \synapticweight_\ranksyn^\postsynaddr$. 
Using this dense representation, the counting defined above becomes:
\begin{equation*}
\mathcal{C}^\postsynaddr(a,t)
= \sum_{\presynaddr, \synapticdelay_\ranksyn^\postsynaddr} \kernel^\postsynaddr(\presynaddr_\ranksyn^\postsynaddr, \synapticdelay_\ranksyn^\postsynaddr) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn^\postsynaddr)
\end{equation*}
This shows that $\mathcal{C}^\postsynaddr$ is a temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}^\postsynaddr = \kernel^\postsynaddr \ast A$.
This well-known computation defines a time-invariant, differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different patterns in the event stream.
In particular, we may extend the convolution to a 3D convolution 
such that the representation would also benefit from spatial invariance. The use of spatiotemporal filters on a stream of events was shown to improve CNN performances for an action recognition task in~\cite{ghosh2022conv3D}.
For that, we design 3D kernels of shape  $(\Kx, \Ky, \Ktime) = (15, 15, 8)$, respectively representing the two spatial dimensions and the range of delays.
The loss function of the MLR model is the binary cross entropy on the output of the classification layer. Simulations are performed thanks to the PyTorch library using gradient descent with Adam ($2^{12}$ epochs and a learning rate of $10^{-5}$). 

\subsection{Task definition: fast motion detection}
%%%-----------------------------------------------------------------
\note{c'est la meme expe faite par baudot et modelisée par kremkow}
Let us now define a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the field of vision. In animals with a fovea, this is particularly useful as it allows to move the area with the highest density of photoreceptors in the environment, for example at a point of interest. For example, saccades are rapid movements of the eyeball that reposition the center of vision. In humans, these are very frequent (on average one every 3 per second~\citep{dandekar_neural_2012}). They are produced very rapidly (about $80~\ms$) and display a whole range of speeds. At a more microscopic scale, the human gaze moves with an incessant drift similar to a Brownian-like trajectory~\citep{poletti_head-eye_2015}. To maintain the full generality of the task, we will define eye movements using a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D motions in polar coordinates. Based on the distribution of biological motions, we simplified it by selecting a set of eye movements as the Cartesian product of N-V-phi=8 linearly spaced motion directions and N-Y-velocity=7 different velocities (see inset of \seeFig{motion_task}-(a)). Note that the velocities are sampled on a geometric scale between $1/4$ and $v-speed_base = 4$. Next, we define one trajectory of gaze as segments whose duration is drawn from a Poisson distribution with an average block length = $20~\ms$ - similar to a Lévy flight~\citep{}. Finally, the trajectory is integrated by assuming 1/ that velocities are uniformly and independently sampled from the set of different motion-sets and 2/ that motion is uniform during a time segment. The resulting instances yield trajectories qualitatively similar to those observed for human eye movements (\seeFig{motion_task}-(a)).
